Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2016-11-11 paper/intro_v3.tex
edits to numerical results, intro

Physical and engineering systems are often described using sophisticated mathematical models, such as coupled, nonlinear partial differential equations (PDEs). The inverse problem seeks to infer unknown model parameters using observed data; often these observations are limited, noisy and provide only indirect information about the unknown parameters.
%~\cite{Taran05, BanksKuhn89}. 
Many such inverse problems present a major computational challenge, since their solution involves repeated solutions of the mathematical model.  In many cases, a given physical system can be represented with varying degrees of fidelity by different models. A higher-fidelity model more accurately represents reality, but is also usually more computationally expensive to solve. Lower-fidelity models reduce the computational cost of the forward problem, at the expense of introducing approximations.
In this work, we formulate an approach for leveraging multiple models of varying fidelities in solution of an inverse problem.

@@

The simultaneous use of multiple models of varying fidelity for forward simulations is well established in some fields. Ref.~\cite{Liuetal03} categorizes two main strategies for combining models: hierarchical and concurrent methods. Hierarchical methods (also known as information-passing or sequential methods) take the results of a simulation using the high-fidelity model and use them to inform a lower-fidelity model that is used globally. An example application is the modeling of the molecular structure of a material to determine parameters for constitutive equations~\cite{Haoetal03,Weietal04}. In contrast, concurrent methods simultaneously solve the higher- and lower-fidelity models in different parts of the domain. Applications include computational mechanics~\cite{Khareetal08,Prudetal08} and fluid dynamics~\cite{vanOpstaletal15,LucKinBer02,FatGerQua01}.
% KW: I feel that we may be missing some important literature here, e.g., the work of Alexander, Tartakovsky
We focus on concurrent methods of combining models, which have desirable features: in the case where the high-fidelity model is nonlinear, replacing it with a linear lower-fidelity model in most of the domain can reduce the need for iterative solves;
% KW: what does 'reduce the need' mean? you still need iterative solves right? 
when the high-fidelity model has a fine resolution and/or many parameters, replacing it with a lower-fidelity model can reduce the number of degrees of freedom of the mixed-fidelity model.

Goal-oriented approaches prioritize accuracy in the QoI over accuracy in the states and/or parameters; one aims to reduce the cost of solving the problem (whether forward or inverse) while maintaining acceptably low error estimate of the QoI. In the context of the forward problem, goal-oriented methods have been developed for mesh refinement \cite{PrudOden99,VendDarm00,BecRann01,Yano12} and for model adaptivity~\cite{OdenPrudetal06,BraackErn03}. 
%where the goal is to optimally mix models differing in their mesh resolution and/or physics included. 
These methods derive an adjoint-based a posteriori error estimate for an output functional, and this estimate is used to guide adaptive mesh and/or model refinement.
% KW: moved Yano reference. please replace reference to PhD thesis by a journal paper:
%@article{Yano2012,
%  title={An optimization-based framework for anisotropic simplex mesh adaptation},
%  author={Yano, Masayuki and Darmofal, David L},
%  journal={Journal of Computational Physics},
%  volume={231},
%  number={22},
%  pages={7626--7649},
%  year={2012},
%  publisher={Elsevier}
%}

>>>>

Physical and engineering systems are often described using sophisticated mathematical models, such as coupled, nonlinear partial differential equations (PDEs). The inverse problem seeks to infer unknown model parameters using observed data; often these observations are limited, noisy and provide only indirect information about the unknown parameters. Many such inverse problems present a major computational challenge, since their solution involves repeated solutions of the mathematical model. In many cases, a given physical system can be represented with varying degrees of fidelity by different models. A higher-fidelity model more accurately represents reality, but is also usually more computationally expensive to solve. Lower-fidelity models reduce the computational cost of the forward problem, at the expense of introducing approximations.
In this work, we formulate an approach for leveraging multiple models of varying fidelities in solution of an inverse problem.

@@

The simultaneous use of multiple models of varying fidelity for forward simulations is well established in some fields. Ref.~\cite{Liuetal03} categorizes two main strategies for combining models: hierarchical and concurrent methods. Hierarchical methods (also known as information-passing or sequential methods) take the results of a simulation using the high-fidelity model and use them to inform a lower-fidelity model that is used globally. An example application is the modeling of the molecular structure of a material to determine parameters for constitutive equations~\cite{Haoetal03,Weietal04}. In contrast, concurrent methods (also called hybrid methods) simultaneously solve the higher- and lower-fidelity models in different parts of the domain. Applications include computational mechanics~\cite{Khareetal08,Prudetal08} and fluid dynamics~\cite{AlexGarTar02,FatGerQua01,Garcetal99,LucKinBer02,vanOpstaletal15,WadErw90}.
% KW: I feel that we may be missing some important literature here, e.g., the work of Alexander, Tartakovsky
We focus on concurrent methods of combining models, which have desirable features: in the case where the high-fidelity model is nonlinear, replacing it with a linear lower-fidelity model in most of the domain can reduce the number of iterative solves needed; when the high-fidelity model has a fine resolution and/or many parameters, replacing it with a lower-fidelity model can reduce the number of degrees of freedom of the mixed-fidelity model.

Goal-oriented approaches prioritize accuracy in the QoI over accuracy in the states and/or parameters; one aims to reduce the cost of solving the problem (whether forward or inverse) while maintaining acceptably low error estimate of the QoI. In the context of the forward problem, goal-oriented methods have been developed for mesh refinement \cite{BecRann01,PrudOden99,VendDarm00,Yano12} and for model adaptivity~\cite{BraackErn03,OdenPrudetal06}. These methods derive an adjoint-based a posteriori error estimate for an output functional, and this estimate is used to guide adaptive mesh and/or model refinement.

### 2016-11-11 paper/numerical_experiments_v3.tex
edits to numerical results, intro

\begin{figure}[htbp]
\centering
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_setup.png}
  \label{subfig:obsSetup}
}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err0.png}
  \label{subfig:obsLF}
}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err1.png}
}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2.png}
  \label{subfig:obsMFlast}
}
  \caption{Effects of increasing the QoI region. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}): the error estimate decompositions for different mixed-fidelity models.}
  \label{fig:qoiStudy}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_setup.png}
  \label{subfig:obsSetup2}
}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err0.png}
  \label{subfig:obsLF2}
}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err1.png}
}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2.png}
  \label{subfig:obsMFlast2}
}
  \caption{Effects of increasing the number of observations. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}): the error estimate decompositions for different mixed-fidelity models.}
  \label{fig:dataStudy}
\end{figure}

@@

% KW: I'm commenting out this next statement, unless you strongly disagree.
%Although using such a pair of models has similarities to the problem of adaptive mesh refinement, we note that in this example only the parameter field changes in its level of refinement, not the state. Should the two models differ in the resolution of the state variables instead of the parameters, it is more efficient to use the approach discussed in \cite{BecVex05}.

@@

We now present the results of solving the inference problem using \cref{alg:refSeries}, with a relative error tolerance of $0.1\%$. At each iteration, we choose the $2\%$ of the basis functions with the largest error for model refinement; since each linear Lagrange basis function has eight elements in its support, the number of additional elements marked for refinement in each iteration is usually greater than for the 2D examples. All simulations are run on a single processor; we use the default nonlinear solver in \texttt{libMesh} \cite{libMeshPaper} (Newton's method with Brent line-search), and linear solves are performed using PETSC's GMRES solver, preconditioned by incomplete factorization. % KW: need to add a citation to papers for libMesh and PETSC?

@@

We notice that, compared to the results in \cref{fig:baseErr}, the error estimates for the low-fidelity and first mixed-fidelity models are far from the true errors. This can be attributed to the linearization about $\Psi_{LF}$ and $\Psi_{MF_{1}}$ instead of $\Psi_{HF}$ in solving the supplementary adjoints; the large differences in the QoI for the LF and MF01 models compared to the high-fidelity QoI indicate that $\Psi_{LF}$ and $\Psi_{MF_{1}}$ are significantly different from $\Psi_{HF}$. Compared to the pair of models in \cref{sec:cdvcdrSetup}, the low- and high-fidelity models are more dissimilar, even though the nonlinear term in the high-fidelity models remain the same. % KW: don't know what last part of this sentence means. what hi-fi models?
This behavior suggests that one should use more than just the error estimate to decide when to stop the adaptive algorithm; a possible additional criterion could be to require that the predicted high-fidelity QoI $I(q_{MF},u_{MF})+\epsilon_i$ for two consecutive iterations agree to within some tolerance. % KW: what behavior? this last sentence doesn't follow from the previous one. and what predicted hi-fi QOI? Do you mean mixed-fi? I'm not sure this is a good idea. I'd probably delete this last sentence.

>>>>

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_qoi/vs_qoi_setup_sidetrim.png}
  \label{subfig:obsSetup}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err0_nobar.png}
  \label{subfig:obsLF}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2_barnorm.png}
  \label{subfig:obsMFlast}
}
  \caption{Effects of increasing the QoI region. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}): the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:qoiStudy}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_data/vs_data_setup_sidetrim.png}
  \label{subfig:obsSetup2}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err0_nobar.png}
  \label{subfig:obsLF2}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2_barnorm.png}
  \label{subfig:obsMFlast2}
}
  \caption{Effects of increasing the number of observations. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}): the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:dataStudy}
\end{figure}

@@

@@

We now present the results of solving the inference problem using \cref{alg:refSeries}, with a relative error tolerance of $0.1\%$. At each iteration, we choose the $2\%$ of the basis functions with the largest error for model refinement; since each linear Lagrange basis function has eight elements in its support, the number of additional elements marked for refinement in each iteration is usually greater than for the 2D examples. All simulations are run on a single processor; we use the default nonlinear solver in \texttt{libMesh} \cite{libMeshPaper} (Newton's method with Brent line-search), and linear solves are performed using PETSc's \cite{petsc-user-ref} GMRES solver, preconditioned by incomplete factorization.

@@

We notice that, compared to the results in \cref{fig:baseErr}, the error estimates for the low-fidelity and first mixed-fidelity models are far from the true errors. This can be attributed to the linearization about $\Psi_{LF}$ and $\Psi_{MF_{1}}$ instead of $\Psi_{HF}$ in solving the supplementary adjoints; the large differences in the QoI for the LF and MF01 models compared to the high-fidelity QoI indicate that $\Psi_{LF}$ and $\Psi_{MF_{1}}$ are significantly different from $\Psi_{HF}$. Compared to the pair of models in \cref{sec:cdvcdrSetup}, the low- and high-fidelity models in this case are more dissimilar, even though in both \cref{sec:cdvcdrSetup} and this case the nonlinear term in the high-fidelity model \cref{eq:cdvcdrHF} and \cref{eq:cdvcdrHF3D} is a quadratic reaction term. % KW: don't know what last part of this sentence means. what hi-fi models?
The possibility of large inaccuracies in the error estimates of low-fidelity and mostly unrefined mixed-fidelity models suggests that one should use more than just the error estimate to decide when to stop the adaptive algorithm; a possible additional criterion could be to require that the estimated high-fidelity QoI ($I(q_{MF},u_{MF})+\epsilon_i$) for two consecutive iterations agree to within some tolerance. % KW: what behavior? this last sentence doesn't follow from the previous one. and what predicted hi-fi QOI? Do you mean mixed-fi? I'm not sure this is a good idea. I'd probably delete this last sentence.
% or just let the reader decide how to deal with possibly inaccurate initial error estimates?
